Dialect: A Compact Line Notation Based on SMILES

### 2022-03-24 manuscript.tex
Unifies page citation. closes #98

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency (herein "The Article").\cite{anderson:1987}
    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
    \item Weininger's 2003 book chapter, revised in 2008 (herein "The Book Chapter").\cite{weininger:2008}
    \item An online manual maintained by Daylight (herein "The Manual").
    \item The Daylight's toolkit, which implements a SMILES reader and writer (herein "The Toolkit").\cite{daylightToolkit}
\end{itemize}

@@

The Manual was disregarded as a ProtoSMILES source because it merely recapitulates content already present in The Paper and The Book Chapter. Removing this source simplifies the construction of ProtoSMILES without changing its definition. A secondary consideration was the possibility that updates to this web page could change the meaning of ProtoSMILES. However web archives indicate that neither styling nor content of the Manual have changed in at least ten years.

@@

The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p. 83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES. The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.

The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p. 33]{weininger:1988} The comma character is explicitly disallowed in Dialect.

The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p. 86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the Book Chapter,\cite[p. 89]{weininger:2008} but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p. 94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare. To support them burdens implementors with little payoff.

ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not. Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided. The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond). "-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p 85]{weininger:2008} This passage implies that aromatic bonds serve no purpose. Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of the language unnecessarily. Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p. 35]{weininger:1988} default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.

It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself. The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p. 34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]". The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen".\cite[p 97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.

ProtoSMILES ascribes ambiguous meaning to detachments. As noted in the Book Chapter, "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p 88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior. ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p 32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.

@@

Dialect also resolves some internal contradictions within ProtoSMILES. The Article reports the only default valence for nitrogen as three (p. 34),\cite[p 34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p 84]{weininger:2008} Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p 32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p 35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p 34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

>>>>

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency ("The Article").\cite{anderson:1987}
    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter").\cite{weininger:2008}
    \item An online manual maintained by Daylight ("The Manual").
    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit").\cite{daylightToolkit}
\end{itemize}

@@

The Manual was disregarded as a ProtoSMILES source because it merely recapitulates content already present in The Paper and The Book Chapter. Removing this source simplifies the construction of ProtoSMILES without changing its definition. A secondary consideration was the possibility that updates to this web page could change the meaning of ProtoSMILES. However, web archives indicate that neither styling nor content of the Manual have changed in at least ten years.

@@

The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.~83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES. The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.

The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.~33]{weininger:1988} The comma character is explicitly disallowed in Dialect.

The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.~86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.~89]{weininger:2008} but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.~94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare. To support them burdens implementors with little payoff.

ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not. Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided. The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond). "-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p.~85]{weininger:2008} This passage implies that aromatic bonds serve no purpose. Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of the language unnecessarily. Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.~35]{weininger:1988} default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.

It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself. The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.~34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]". The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen".\cite[p.~97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.

ProtoSMILES ascribes ambiguous meaning to detachments. As noted in the Book Chapter, "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p.~88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior. ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p.~32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.

@@

Dialect also resolves some internal contradictions within ProtoSMILES. The Article reports the only default valence for nitrogen as three (p. 34),\cite[p.~34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p.~84]{weininger:2008} Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p.~32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p.~35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p.~34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
